% Blueprint for the Verified VJP Proofs suite (LeanMlir/Proofs/)
%
% Every axiom, definition, and theorem in the suite is annotated below so
% leanblueprint can render the dependency graph. Each block cites its
% Lean declaration via \lean{...}, marks proved claims with \leanok, and
% lists upstream dependencies via \uses{...}.

\chapter{Introduction}
\label{chap:introduction}

On the next-to-last page of my first book, I wrote that I dreamed of a
future where we could take our code and compile it for whatever backend
we desire, via MLIR. I meant it as a wishlist item --- something I expected
to happen eventually, but not soon enough to matter for that work.

It turned out ``eventually'' was about five years.

This is the book I wanted to write back then, and couldn't. The tools
weren't ready. Lean 4 was still in development. MLIR was a research
project. IREE couldn't compile a training step. Now all three are
production-quality, and the thing I wished for is not only possible ---
it's sitting in a GitHub repo, training ResNets on a single GPU, with
machine-checked proofs that every gradient is correct.

\section*{What this book is about}

This book teaches you how backpropagation actually works.

Not the hand-wave version --- ``compute gradients, update weights,
repeat.'' You already know that version. If you've trained a model,
you've called \texttt{.backward()} a thousand times. What you probably
haven't done is open the hood and look at what \texttt{.backward()} is
actually computing, layer by layer, and verify that the computation
is mathematically correct.

That's what we're going to do. For every layer in the modern stack ---
dense, ReLU, convolution, pooling, batch normalization, residual
connections, depthwise convolution, squeeze-and-excitation, layer
normalization, and self-attention --- we will:

\begin{enumerate}
  \item \textbf{Derive the backward pass} from the forward pass, using nothing
    but the chain rule and some index algebra.
  \item \textbf{State the result formally} in Lean 4, a programming language
    that doubles as a theorem prover.
  \item \textbf{Show the MLIR} that compiles to GPU bytecode, and verify it
    matches the math.
  \item \textbf{Train the model} on real data, end-to-end, on your hardware.
\end{enumerate}

Three views of the same computation: the math, the proof, the code
that runs. They agree, and the entire book is about understanding why.

\section*{The thesis}

Here's the punchline, stated up front so you know where we're headed.

Backpropagation is \textbf{one operation} (the vector-Jacobian product),
composed using \textbf{three rules}:

\begin{enumerate}
  \item \textbf{The chain rule} --- if you compose two functions, you compose
    their backward passes.
  \item \textbf{Additive fan-in} --- if two paths feed into a sum (like a
    residual connection), their gradients add.
  \item \textbf{Multiplicative fan-in} --- if two paths feed into a product
    (like squeeze-and-excitation or attention), the product rule
    gives you two backward contributions that add.
\end{enumerate}

And \textbf{five structural tricks} for layers whose Jacobians are dense
but exploitably structured:

\begin{enumerate}
  \item \textbf{Diagonal} --- elementwise activations (ReLU, GELU, Swish).
    The Jacobian is diagonal; the backward is one multiply.
  \item \textbf{Sparse Toeplitz} --- convolutions. The Jacobian is block-sparse
    with a sliding-window structure; the backward is another
    convolution with a reversed kernel.
  \item \textbf{Binary selection} --- max pooling. The Jacobian routes the
    gradient to whichever input was the max.
  \item \textbf{Rank-1 correction to diagonal} --- softmax, batch norm, layer
    norm. The Jacobian is dense but has a clean closed form because
    it's a diagonal plus a rank-1 outer product.
  \item \textbf{Outer product reductions} --- dense layers and matmul. The
    weight gradient is the outer product of the input with the
    output gradient.
\end{enumerate}

That's it. Three rules, five tricks. Every architecture in this book ---
MLP, CNN, ResNet, MobileNet, EfficientNet, Vision Transformer --- and
every architecture in the bestiary --- U-Net, DETR, Mamba, GAN, VAE,
diffusion models, CLIP, GPT-style decoders --- decomposes into some
combination of these eight things. There is no ninth.

I didn't set out to prove this. I set out to train a ResNet with
verified gradients, and by the time I'd done it for enough architectures,
the pattern became impossible to ignore. The framework kept working
without needing new primitives. Once you see it, you can't unsee it,
and the rest of the book is about making sure you see it.

\section*{Who this book is for}

This book assumes you've trained a neural network. Not a fancy one.
If you've fine-tuned a pretrained model, debugged a NaN loss, or
watched a model overfit and wondered why, you're the audience.

I'm going to assume you know what a derivative is, what a matrix is,
and roughly what a convolutional layer does. I'm not going to assume
you know Lean, MLIR, formal verification, or compiler theory. The Lean
code appears throughout the book as a precise specification and a
credibility instrument --- not as a barrier. If you skip every Lean
snippet, you still get a complete book on the mechanics of
backpropagation. The Lean is there for the reader who wants to
\emph{check} me, not the reader who needs to learn formal methods.

This is not a beginner's book. If you're looking for your first
introduction to neural networks, there are several good ones: my
first book is one option, fast.ai's free course is another, and there
are good resources all over the web. Come back to this one when
you've trained something and want to understand what you did.

This is also not a research textbook. I'm not trying to write Goodfellow
or Bishop. The proofs here are for credibility, not for theoretical depth.
If you want the full formal story, Mathlib is open-source and you can
extend it. What this book gives you that a research textbook doesn't is
the bridge from the math to the actual GPU code, and the assurance
that they agree.

If you're me from ten years ago --- interested in ML, willing to put in
the work, not sure where the math stops and the folklore begins --- this
is the book I wish I'd had.

\section*{Why now}

Three things had to happen before this book was possible, and all three
happened in the last few years:

\textbf{Lean 4 reached maturity.} Lean is a programming language that is
also a theorem prover. You can write a function in Lean and then write
a proof that the function does what you claim. If the proof compiles,
the claim is correct --- not because you trust the author, but because
the compiler checked it. Lean 4 (released 2023) is the version that
made this practical for non-specialists: it has a real package manager,
real tooling, a growing math library (Mathlib), and an active community.

\textbf{MLIR and StableHLO stabilized.} MLIR is a compiler infrastructure
for machine learning. StableHLO is a portable operation set built on
MLIR that describes the computations neural networks perform. Together,
they let you specify a training step as a sequence of mathematical
operations and compile it to any hardware backend --- CPU, CUDA, ROCm ---
without writing backend-specific code. This is the layer that turns
``math on paper'' into ``code on the GPU.''

\textbf{IREE became production-ready.} IREE (Intermediate Representation
Execution Environment) is the runtime that takes a compiled StableHLO
module and executes it on actual hardware. It handles memory management,
device allocation, and kernel dispatch. It's what lets us go from a
Lean spec file to a running training loop with no Python anywhere in
the pipeline.

The combination of these three tools means, for the first time, that
a single person can: (a) write a neural network spec in a language
with a type system strong enough to state and check mathematical
properties; (b) compile that spec to GPU-runnable code via a standard
intermediate representation; and (c) train the network end-to-end on
consumer hardware, with formal guarantees about the gradient
computation. That's the pipeline this book is built around.

\section*{Why image recognition}

Same reason as the first book: it's the oldest, most well-understood
corner of deep learning, which means we can introduce primitives one
at a time in a logical order.

We start with MNIST (digits, $28 \times 28$ grayscale) and work up through
CIFAR-10 (color, $32 \times 32$) to Imagenette (real photos, $224 \times 224$).
The architectures form a natural progression:
MLP $\to$ CNN $\to$ ResNet $\to$ MobileNet $\to$ EfficientNet $\to$ ViT.
Each one adds exactly one new structural primitive to the framework. By
the time you reach ViT, you've seen every primitive there is to see for
modern vision --- and as it turns out, for modern language models too,
since transformers are the same architecture in both domains.

The bestiary at the end demonstrates that the same primitives cover
detection (DETR), segmentation (U-Net), generation (diffusion models),
self-supervised learning (MAE, CLIP), sequence modeling (Mamba), and
multi-modal models (Flamingo). Image recognition is the on-ramp; the
destination is the whole modern stack.

\section*{Why Lean}

I considered several options for the ``formal'' half of the book ---
Coq, Agda, Isabelle, or just doing the proofs on paper with LaTeX.
I chose Lean for a practical reason: it reads like code.

If you've trained a model in PyTorch, you can read:

\begin{verbatim}
def dense {m n : Nat} (W : Mat m n) (b : Vec n) (x : Vec m) : Vec n :=
  fun j => finSum m (fun i => x i * W i j) + b j
\end{verbatim}

and understand it. It's a function with typed arguments.
\texttt{\{m n : Nat\}} are the dimensions. \texttt{fun j =>} is a lambda.
\texttt{finSum} is a for loop. You don't need to learn a new paradigm
--- you need to learn maybe twenty keywords and some notation, and then
Lean reads like any other typed functional language.

The other reason is that Lean enforces honesty. If I claim that the
backward pass of batch normalization is a specific three-term formula,
the Lean compiler requires me to prove it. If the proof compiles,
the claim is correct --- not because you trust me, but because the
type checker verified it. There is no hidden hand-waving. Every VJP
in this book --- dense, convolution, batch norm, residual, attention,
all of them --- builds with zero \texttt{sorry}s. The compiler is the
referee.

\section*{How to read this book}

Every chapter after this one follows the same pattern:

\begin{enumerate}
  \item A new architecture (the ``what'')
  \item A new structural primitive (the ``why'')
  \item Derivation of the backward pass (by hand, with indices)
  \item Lean proof (the formal statement)
  \item MLIR snippet (the GPU code)
  \item Training results (proof it works)
\end{enumerate}

You can read straight through, or you can skip to any chapter whose
architecture interests you --- each one is mostly self-contained, with
back-references to earlier chapters for the primitives it builds on.

The Lean code and MLIR snippets are available in the lean4-mlir
repository on GitHub. Every architecture in the book can be trained
on your hardware. Appendix A walks you through the toolchain setup;
if you're impatient, there's a Docker image that gets you from zero
to training MNIST in one command.

\section*{How this book is organized}

\textbf{Basics.} We build the framework from scratch using simple
architectures:

\begin{itemize}
  \item \textbf{Chapter 2: Tensor Calculus} --- Axioms.
  \item \textbf{Chapter 3: MNIST 1D MLP} --- Dense layers, ReLU, softmax CE,
    the chain rule. The most important chapter: everything after is
    adding more layers.
  \item \textbf{Chapter 4: MNIST 2D CNN} --- Convolution, max pooling. The
    backward pass of conv is another conv with a reversed kernel.
  \item \textbf{Chapter 5: CIFAR with BatchNorm} --- The first dense Jacobian
    with a clean closed form. Three terms, one cancellation, and you
    never have to derive it again.
  \item \textbf{Chapter 6: ResNet-34} --- Residual connections. Gradients from
    two paths add at the input.
\end{itemize}

\textbf{Spanning models.} We walk up the architecture ladder, one new
primitive per chapter:

\begin{itemize}
  \item \textbf{Chapter 7: MobileNetV2} --- Depthwise convolution. Same as
    regular conv, but diagonal in the channel dimension.
  \item \textbf{Chapter 8: EfficientNet} --- Squeeze-and-excitation. The
    product rule: main path times gate, gradients from both paths add.
  \item \textbf{Chapter 9: LayerNorm and GELU} --- Batchnorm in disguise.
    Activation functions.
  \item \textbf{Chapter 10: Vision Transformer} --- Self-attention. Softmax
    in the middle of the network, three-way fan-in at the input. The
    capstone.
\end{itemize}

\textbf{The bestiary.} A catalog of $\sim$20 additional architectures,
from U-Net to Mamba to diffusion models, each decomposed into the
framework's primitives in $\sim$2 pages.

\begin{itemize}
  \item \textbf{Chapter 11: The Bestiary} --- Organized by the question
    you're asking: ``how do I do detection?'', ``how do I do generation?'',
    etc. Plus a micro-design appendix for decoding the latest LLM
    training tricks.
\end{itemize}

\textbf{Appendices.} Toolchain setup (A) and the complete proof framework
as a reference (B).

\section*{A note on the proofs}

The proofs in this book compile with \textbf{zero \texttt{sorry}s}. Every
VJP correctness theorem --- dense layers, convolution, batch normalization,
residual connections, depthwise convolution, squeeze-and-excitation,
layer normalization, and self-attention --- is machine-checked by Lean's
type system. If it builds, it's correct.

The full axiom list, grouped by role, is in Appendix~\ref{app:axioms}.

Everything else is proved from these axioms: the VJP composition
rule (chain rule), additive and multiplicative fan-in, and the
correctness of every layer's backward pass. The proofs are not deep
--- they're undergraduate calculus, formalized. The value is not that
the math is surprising, but that the compiler verified it matches
the exact MLIR code that runs on the GPU.

\section*{Let's go}

In 2018, I was on a small team that won Stanford's DAWNBench
competition --- training ImageNet in 3 hours for \$25 on commodity
hardware. The lesson from that experience was simple: most of the
apparent difficulty in modern deep learning isn't fundamental. It's
accumulated folklore, missing tools, and a culture that takes the math
on faith.

This book applies that same lesson to the math itself. The folklore
says backpropagation is complicated. It isn't. It's three rules, five
tricks, and a lot of bookkeeping. The tools to verify this claim
finally exist, and you're holding the result.

Three of my favorite citations of the first book are from high school
students. I don't think that's an accident. The barriers to doing real
ML research used to be infrastructure --- compute, data, institutional
access. Those barriers have mostly fallen. What remains is the
smaller barrier of understanding what you're doing at the level where
you can verify it, not just run it.

This book is for the person who wants to cross that barrier. It might
be you. It might be the next kid who picks it up in a library somewhere
and decides to see how far they can go.

Let's find out.

\chapter{Tensor: Calculus Foundations}
\label{chap:tensor}

The algebraic framework: 1D \(\mathsf{Vec}\), 2D \(\mathsf{Mat}\), 3D
\(\mathsf{Tensor3}\) types, the partial-derivative function \(\pdiv\) and
its structural rules, and three VJP record types (\(\HasVJP\),
\(\HasVJPMat\), \(\HasVJPthree\)) that package a backward function together
with its correctness claim.

\begin{axiom}[Partial derivative]
  \label{ax:pdiv}
  \lean{Proofs.pdiv}
  \leanok
  The partial derivative function. For \(f : \Vect{m} \to \Vect{n}\),
  \(\pdiv\, f\, x\, i\, j\) is the \((i, j)\)-entry of the Jacobian at \(x\).
\end{axiom}

\begin{axiom}[Chain rule]
  \label{ax:pdiv_comp}
  \lean{Proofs.pdiv_comp}
  \leanok
  \uses{ax:pdiv}
  \(\pdiv(g \circ f)\, x\, i\, k = \sum_j \pdiv f\, x\, i\, j \cdot \pdiv g\, (f\, x)\, j\, k\).
\end{axiom}

\begin{axiom}[Sum rule]
  \label{ax:pdiv_add}
  \lean{Proofs.pdiv_add}
  \leanok
  \uses{ax:pdiv}
  Linearity of the derivative.
\end{axiom}

\begin{axiom}[Product rule]
  \label{ax:pdiv_mul}
  \lean{Proofs.pdiv_mul}
  \leanok
  \uses{ax:pdiv}
  Pointwise Leibniz rule for elementwise products.
\end{axiom}

\begin{axiom}[Identity Jacobian]
  \label{ax:pdiv_id}
  \lean{Proofs.pdiv_id}
  \leanok
  \uses{ax:pdiv}
  \(\pdiv(\mathrm{id})\, x\, i\, j = \delta_{ij}\).
\end{axiom}

\begin{axiom}[Constant has zero Jacobian]
  \label{ax:pdiv_const}
  \lean{Proofs.pdiv_const}
  \leanok
  \uses{ax:pdiv}
\end{axiom}

\begin{axiom}[Gather / reindex Jacobian]
  \label{ax:pdiv_reindex}
  \lean{Proofs.pdiv_reindex}
  \leanok
  \uses{ax:pdiv}
  Covers permutations, reshapes, slicing. Generalizes \texttt{pdiv\_id}.
\end{axiom}

\begin{axiom}[Row-independence for matrices]
  \label{ax:pdivMat_rowIndep}
  \lean{Proofs.pdivMat_rowIndep}
  \leanok
  \uses{ax:pdiv}
  Applying a vector function per-row to a matrix has block-diagonal Jacobian.
  The one genuinely-new primitive at the \(\mathsf{Mat}\) level.
\end{axiom}

\begin{theorem}[Finite-sum rule]
  \label{thm:pdiv_finset_sum}
  \lean{Proofs.pdiv_finset_sum}
  \leanok
  \uses{ax:pdiv, ax:pdiv_add, ax:pdiv_const}
  Linearity extended to arbitrary finite sums by induction.
\end{theorem}

\begin{theorem}[VJP chain rule]
  \label{thm:vjp_comp}
  \lean{Proofs.vjp_comp}
  \leanok
  \uses{ax:pdiv, ax:pdiv_comp}
  Given \(\HasVJP\, f\) and \(\HasVJP\, g\), get \(\HasVJP\, (g \circ f)\).
\end{theorem}

\begin{theorem}[Additive fan-in VJP]
  \label{thm:biPath_has_vjp}
  \lean{Proofs.biPath_has_vjp}
  \leanok
  \uses{ax:pdiv, ax:pdiv_add}
  VJP of \(f + g\). Used for residual connections.
\end{theorem}

\begin{theorem}[Multiplicative fan-in VJP]
  \label{thm:elemwiseProduct_has_vjp}
  \lean{Proofs.elemwiseProduct_has_vjp}
  \leanok
  \uses{ax:pdiv, ax:pdiv_mul}
  VJP of elementwise product. Used for Squeeze-and-Excitation.
\end{theorem}

\begin{theorem}[Identity VJP]
  \label{thm:identity_has_vjp}
  \lean{Proofs.identity_has_vjp}
  \leanok
  \uses{ax:pdiv, ax:pdiv_id}
\end{theorem}

\begin{theorem}[Matrix-level chain rule]
  \label{thm:vjpMat_comp}
  \lean{Proofs.vjpMat_comp}
  \leanok
  \uses{thm:vjp_comp}
  Lifts \texttt{vjp\_comp} via the \(\mathsf{Mat.flatten}\) bijection.
\end{theorem}

\begin{theorem}[Matrix-level additive fan-in]
  \label{thm:biPathMat_has_vjp}
  \lean{Proofs.biPathMat_has_vjp}
  \leanok
  \uses{thm:biPath_has_vjp}
\end{theorem}

\begin{theorem}[Matrix-level identity]
  \label{thm:identityMat_has_vjp}
  \lean{Proofs.identityMat_has_vjp}
  \leanok
  \uses{thm:identity_has_vjp}
\end{theorem}

\begin{theorem}[Scalar-scale Jacobian]
  \label{thm:pdivMat_scalarScale}
  \lean{Proofs.pdivMat_scalarScale}
  \leanok
  \uses{ax:pdiv_mul, ax:pdiv_const, ax:pdiv_id}
  Phase 6 derivation.
\end{theorem}

\begin{theorem}[Transpose Jacobian]
  \label{thm:pdivMat_transpose}
  \lean{Proofs.pdivMat_transpose}
  \leanok
  \uses{ax:pdiv_reindex}
\end{theorem}

\begin{theorem}[Matmul Jacobian, left factor fixed]
  \label{thm:pdivMat_matmul_left_const}
  \lean{Proofs.pdivMat_matmul_left_const}
  \leanok
  \uses{ax:pdiv_const, ax:pdiv_reindex, thm:pdiv_finset_sum}
  Phase 6: derived, not axiomatized.
\end{theorem}

\begin{theorem}[Matmul Jacobian, right factor fixed]
  \label{thm:pdivMat_matmul_right_const}
  \lean{Proofs.pdivMat_matmul_right_const}
  \leanok
  \uses{ax:pdiv_const, ax:pdiv_reindex, thm:pdiv_finset_sum}
\end{theorem}

\begin{theorem}[Matmul VJP, left factor fixed]
  \label{thm:matmul_left_const_has_vjp}
  \lean{Proofs.matmul_left_const_has_vjp}
  \leanok
  \uses{thm:pdivMat_matmul_left_const}
\end{theorem}

\begin{theorem}[Matmul VJP, right factor fixed]
  \label{thm:matmul_right_const_has_vjp}
  \lean{Proofs.matmul_right_const_has_vjp}
  \leanok
  \uses{thm:pdivMat_matmul_right_const}
\end{theorem}

\begin{theorem}[Scalar-scale VJP]
  \label{thm:scalarScale_has_vjp}
  \lean{Proofs.scalarScale_has_vjp}
  \leanok
  \uses{thm:pdivMat_scalarScale}
\end{theorem}

\begin{theorem}[Transpose VJP]
  \label{thm:transpose_has_vjp}
  \lean{Proofs.transpose_has_vjp}
  \leanok
  \uses{thm:pdivMat_transpose}
\end{theorem}

\begin{theorem}[Row-wise lifting of any \(\HasVJP\)]
  \label{thm:rowwise_has_vjp_mat}
  \lean{Proofs.rowwise_has_vjp_mat}
  \leanok
  \uses{ax:pdivMat_rowIndep}
  Phase 8 generic helper: lifts any \(\HasVJP\, (g : \Vect{n} \to \Vect{p})\)
  to a \(\HasVJPMat\) on \(\Mat{m}{n} \to \Mat{m}{p}\).
\end{theorem}

\begin{theorem}[3D chain rule]
  \label{thm:pdiv3_comp}
  \lean{Proofs.pdiv3_comp}
  \leanok
  \uses{ax:pdiv_comp}
\end{theorem}

\begin{theorem}[3D VJP chain rule]
  \label{thm:vjp3_comp}
  \lean{Proofs.vjp3_comp}
  \leanok
  \uses{thm:pdiv3_comp}
\end{theorem}

\begin{theorem}[3D additive fan-in]
  \label{thm:biPath3_has_vjp}
  \lean{Proofs.biPath3_has_vjp}
  \leanok
  \uses{ax:pdiv_add}
\end{theorem}

\chapter{MLP: Dense Layer}
\label{chap:mlp}

The workhorse: \(y = Wx + b\) plus ReLU and softmax cross-entropy loss.

\begin{axiom}[Dense Jacobian wrt input]
  \label{ax:pdiv_dense}
  \lean{Proofs.pdiv_dense}
  \leanok
  \uses{ax:pdiv}
  \(\partial (Wx + b)_j / \partial x_i = W_{ij}\).
\end{axiom}

\begin{axiom}[Dense Jacobian wrt weight]
  \label{ax:pdiv_dense_W}
  \lean{Proofs.pdiv_dense_W}
  \leanok
  \uses{ax:pdiv}
  \(\partial (Wx + b)_j / \partial W_{i'j'} = x_{i'} \delta_{jj'}\). Phase 7.
\end{axiom}

\begin{axiom}[ReLU Jacobian]
  \label{ax:pdiv_relu}
  \lean{Proofs.pdiv_relu}
  \leanok
  \uses{ax:pdiv}
\end{axiom}

\begin{axiom}[Softmax cross-entropy gradient]
  \label{ax:softmaxCE_grad}
  \lean{Proofs.softmaxCE_grad}
  \leanok
  \uses{ax:pdiv}
  \(\partial L / \partial z = \mathrm{softmax}(z) - \mathrm{onehot}(y)\).
\end{axiom}

\begin{theorem}[Dense VJP]
  \label{thm:dense_has_vjp}
  \lean{Proofs.dense_has_vjp}
  \leanok
  \uses{ax:pdiv_dense}
\end{theorem}

\begin{theorem}[Dense weight gradient is the outer product]
  \label{thm:dense_weight_grad_correct}
  \lean{Proofs.dense_weight_grad_correct}
  \leanok
  \uses{ax:pdiv_dense_W}
  \(dW = x \otimes dy\). Phase 7 promoted from vacuous \texttt{rfl} to theorem.
\end{theorem}

\begin{theorem}[Dense bias gradient is identity]
  \label{thm:dense_bias_grad_correct}
  \lean{Proofs.dense_bias_grad_correct}
  \leanok
  \uses{ax:pdiv_add, ax:pdiv_const, ax:pdiv_id}
  \(db = dy\). Phase 7: derived, no new axiom.
\end{theorem}

\begin{theorem}[ReLU VJP]
  \label{thm:relu_has_vjp}
  \lean{Proofs.relu_has_vjp}
  \leanok
  \uses{ax:pdiv_relu}
\end{theorem}

\chapter{CNN: Convolution and Pooling}
\label{chap:cnn}

\begin{axiom}[Conv2d forward]
  \label{ax:conv2d}
  \lean{Proofs.conv2d}
  \leanok
  Opaque function declaration. Codegen provides the actual
  \texttt{stablehlo.convolution}; proofs reason about it as a black box.
\end{axiom}

\begin{axiom}[Conv2d input VJP]
  \label{ax:conv2d_has_vjp3}
  \lean{Proofs.conv2d_has_vjp3}
  \leanok
  \uses{ax:conv2d}
  Reversed-kernel convolution formula, numerically gradient-checked.
\end{axiom}

\begin{axiom}[Conv2d weight VJP]
  \label{ax:conv2d_weight_grad_has_vjp}
  \lean{Proofs.conv2d_weight_grad_has_vjp}
  \leanok
  \uses{ax:conv2d}
  Phase 7: the transpose-trick formula via \texttt{Kernel4.flatten}.
\end{axiom}

\begin{axiom}[Conv2d bias VJP]
  \label{ax:conv2d_bias_grad_has_vjp}
  \lean{Proofs.conv2d_bias_grad_has_vjp}
  \leanok
  \uses{ax:conv2d}
  Phase 9: sum cotangent over spatial dims per channel.
\end{axiom}

\begin{axiom}[MaxPool 2x2 stride 2 forward]
  \label{ax:maxPool2}
  \lean{Proofs.maxPool2}
  \leanok
\end{axiom}

\begin{axiom}[MaxPool input VJP]
  \label{ax:maxPool2_has_vjp3}
  \lean{Proofs.maxPool2_has_vjp3}
  \leanok
  \uses{ax:maxPool2}
  Routes gradient to the argmax position within each pooling window.
\end{axiom}

\chapter{BatchNorm: the hard one}
\label{chap:bn}

\begin{axiom}[BN affine step Jacobian]
  \label{ax:pdiv_bnAffine}
  \lean{Proofs.pdiv_bnAffine}
  \leanok
  \uses{ax:pdiv}
  \(\partial (\gamma v + \beta) / \partial v_i = \gamma \delta_{ij}\).
\end{axiom}

\begin{axiom}[BN centering Jacobian]
  \label{ax:pdiv_bnCentered}
  \lean{Proofs.pdiv_bnCentered}
  \leanok
  \uses{ax:pdiv}
  \(\partial (x_j - \mu) / \partial x_i = \delta_{ij} - 1/n\).
\end{axiom}

\begin{axiom}[BN inverse-stddev broadcast Jacobian]
  \label{ax:pdiv_bnIstdBroadcast}
  \lean{Proofs.pdiv_bnIstdBroadcast}
  \leanok
  \uses{ax:pdiv}
\end{axiom}

\begin{theorem}[BN normalize 3-term VJP]
  \label{thm:bnNormalize_has_vjp}
  \lean{Proofs.bnNormalize_has_vjp}
  \leanok
  \uses{ax:pdiv_bnCentered, ax:pdiv_bnIstdBroadcast, ax:pdiv_mul}
  The consolidated 3-term backward: factor \(\mathrm{bnXhat}\) as
  \((x - \mu) \cdot \mathrm{istd}\), apply product rule, collapse.
\end{theorem}

\begin{theorem}[BN affine VJP]
  \label{thm:bnAffine_has_vjp}
  \lean{Proofs.bnAffine_has_vjp}
  \leanok
  \uses{ax:pdiv_bnAffine}
\end{theorem}

\begin{theorem}[Full BN VJP]
  \label{thm:bn_has_vjp}
  \lean{Proofs.bn_has_vjp}
  \leanok
  \uses{thm:bnNormalize_has_vjp, thm:bnAffine_has_vjp, thm:vjp_comp}
  \(\mathrm{bn} = \mathrm{affine} \circ \mathrm{normalize}\).
\end{theorem}

\chapter{Residual: Skip Connections}
\label{chap:residual}

\begin{theorem}[Residual block VJP]
  \label{thm:residual_has_vjp}
  \lean{Proofs.residual_has_vjp}
  \leanok
  \uses{thm:biPath_has_vjp, thm:identity_has_vjp}
  Residual = \(f + \mathrm{id}\).
\end{theorem}

\chapter{Depthwise Convolution}
\label{chap:depthwise}

\begin{axiom}[Depthwise conv forward]
  \label{ax:depthwiseConv2d}
  \lean{Proofs.depthwiseConv2d}
  \leanok
\end{axiom}

\begin{axiom}[Depthwise input VJP]
  \label{ax:depthwise_has_vjp3}
  \lean{Proofs.depthwise_has_vjp3}
  \leanok
  \uses{ax:depthwiseConv2d}
\end{axiom}

\begin{axiom}[Depthwise weight VJP]
  \label{ax:depthwise_weight_grad_has_vjp3}
  \lean{Proofs.depthwise_weight_grad_has_vjp3}
  \leanok
  \uses{ax:depthwiseConv2d}
  Phase 7: reuses \(\HasVJPthree\) directly since \texttt{DepthwiseKernel}
  is definitionally \(\mathsf{Tensor3}\).
\end{axiom}

\begin{axiom}[Depthwise bias VJP]
  \label{ax:depthwise_bias_grad_has_vjp}
  \lean{Proofs.depthwise_bias_grad_has_vjp}
  \leanok
  \uses{ax:depthwiseConv2d}
  Phase 9.
\end{axiom}

\chapter{Squeeze-and-Excitation}
\label{chap:se}

\begin{theorem}[SE block VJP]
  \label{thm:seBlock_has_vjp}
  \lean{Proofs.seBlock_has_vjp}
  \leanok
  \uses{thm:elemwiseProduct_has_vjp, thm:dense_has_vjp, thm:identity_has_vjp}
  SE multiplies input by a sigmoid-gated channel mask.
\end{theorem}

\chapter{LayerNorm and GELU}
\label{chap:layernorm}

\begin{axiom}[GELU scalar function]
  \label{ax:geluScalar}
  \lean{Proofs.geluScalar}
  \leanok
\end{axiom}

\begin{axiom}[GELU scalar derivative]
  \label{ax:geluScalarDeriv}
  \lean{Proofs.geluScalarDeriv}
  \leanok
\end{axiom}

\begin{axiom}[GELU Jacobian]
  \label{ax:pdiv_gelu}
  \lean{Proofs.pdiv_gelu}
  \leanok
  \uses{ax:pdiv, ax:geluScalarDeriv}
  Diagonal activation Jacobian.
\end{axiom}

\begin{theorem}[LayerNorm VJP]
  \label{thm:layerNorm_has_vjp}
  \lean{Proofs.layerNorm_has_vjp}
  \leanok
  \uses{thm:bn_has_vjp}
  LayerNorm = BatchNorm on a different axis; same primitive.
\end{theorem}

\begin{theorem}[GELU VJP]
  \label{thm:gelu_has_vjp}
  \lean{Proofs.gelu_has_vjp}
  \leanok
  \uses{ax:pdiv_gelu}
\end{theorem}

\chapter{Attention and the ViT finale}
\label{chap:attention}

\begin{axiom}[Softmax Jacobian]
  \label{ax:pdiv_softmax}
  \lean{Proofs.pdiv_softmax}
  \leanok
  \uses{ax:pdiv}
  \(\partial \mathrm{softmax}(z)_j / \partial z_i = p_j (\delta_{ij} - p_i)\).
\end{axiom}

\begin{theorem}[Standalone softmax VJP]
  \label{thm:softmax_has_vjp}
  \lean{Proofs.softmax_has_vjp}
  \leanok
  \uses{ax:pdiv_softmax}
  Closed-form collapse: \(dz_i = p_i (dy_i - \langle p, dy \rangle)\).
\end{theorem}

\begin{theorem}[Row-wise softmax VJP on a matrix]
  \label{thm:rowSoftmax_has_vjp_mat}
  \lean{Proofs.rowSoftmax_has_vjp_mat}
  \leanok
  \uses{ax:pdivMat_rowIndep, thm:softmax_has_vjp}
\end{theorem}

\begin{theorem}[SDPA backward wrt Q]
  \label{thm:sdpa_back_Q_correct}
  \lean{Proofs.sdpa_back_Q_correct}
  \leanok
  \uses{thm:matmul_right_const_has_vjp, thm:scalarScale_has_vjp, thm:rowSoftmax_has_vjp_mat, thm:vjpMat_comp}
  Composed from four proved \(\HasVJPMat\) building blocks.
\end{theorem}

\begin{theorem}[SDPA backward wrt K]
  \label{thm:sdpa_back_K_correct}
  \lean{Proofs.sdpa_back_K_correct}
  \leanok
  \uses{thm:matmul_right_const_has_vjp, thm:matmul_left_const_has_vjp, thm:scalarScale_has_vjp, thm:rowSoftmax_has_vjp_mat, thm:transpose_has_vjp, thm:vjpMat_comp}
\end{theorem}

\begin{theorem}[SDPA backward wrt V]
  \label{thm:sdpa_back_V_correct}
  \lean{Proofs.sdpa_back_V_correct}
  \leanok
  \uses{thm:matmul_left_const_has_vjp}
\end{theorem}

\begin{axiom}[Multi-head SDPA VJP]
  \label{ax:mhsa_has_vjp_mat}
  \lean{Proofs.mhsa_has_vjp_mat}
  \leanok
  Phase 8: the one bundled axiom needed to lift single-head SDPA to full
  multi-head attention. The "vmap over head axis" primitive.
\end{axiom}

\begin{theorem}[Per-token LayerNorm lifted to a matrix]
  \label{thm:layerNorm_per_token_has_vjp_mat}
  \lean{Proofs.layerNorm_per_token_has_vjp_mat}
  \leanok
  \uses{thm:rowwise_has_vjp_mat, thm:layerNorm_has_vjp}
\end{theorem}

\begin{theorem}[Per-token dense lifted to a matrix]
  \label{thm:dense_per_token_has_vjp_mat}
  \lean{Proofs.dense_per_token_has_vjp_mat}
  \leanok
  \uses{thm:rowwise_has_vjp_mat, thm:dense_has_vjp}
\end{theorem}

\begin{theorem}[Per-token GELU lifted to a matrix]
  \label{thm:gelu_per_token_has_vjp_mat}
  \lean{Proofs.gelu_per_token_has_vjp_mat}
  \leanok
  \uses{thm:rowwise_has_vjp_mat, thm:gelu_has_vjp}
\end{theorem}

\begin{theorem}[Transformer MLP sublayer VJP]
  \label{thm:transformerMlp_has_vjp_mat}
  \lean{Proofs.transformerMlp_has_vjp_mat}
  \leanok
  \uses{thm:dense_per_token_has_vjp_mat, thm:gelu_per_token_has_vjp_mat, thm:vjpMat_comp}
\end{theorem}

\begin{theorem}[Transformer attention sublayer with residual VJP]
  \label{thm:transformerAttnSublayer_has_vjp_mat}
  \lean{Proofs.transformerAttnSublayer_has_vjp_mat}
  \leanok
  \uses{thm:biPathMat_has_vjp, thm:identityMat_has_vjp, thm:layerNorm_per_token_has_vjp_mat, ax:mhsa_has_vjp_mat, thm:vjpMat_comp}
\end{theorem}

\begin{theorem}[Transformer MLP sublayer with residual VJP]
  \label{thm:transformerMlpSublayer_has_vjp_mat}
  \lean{Proofs.transformerMlpSublayer_has_vjp_mat}
  \leanok
  \uses{thm:biPathMat_has_vjp, thm:identityMat_has_vjp, thm:layerNorm_per_token_has_vjp_mat, thm:transformerMlp_has_vjp_mat, thm:vjpMat_comp}
\end{theorem}

\begin{theorem}[Transformer block VJP]
  \label{thm:transformerBlock_has_vjp_mat}
  \lean{Proofs.transformerBlock_has_vjp_mat}
  \leanok
  \uses{thm:transformerAttnSublayer_has_vjp_mat, thm:transformerMlpSublayer_has_vjp_mat, thm:vjpMat_comp}
  One \texttt{vjpMat\_comp} glueing the two sublayers.
\end{theorem}

\begin{theorem}[Transformer tower, any depth via induction]
  \label{thm:transformerTower_has_vjp_mat}
  \lean{Proofs.transformerTower_has_vjp_mat}
  \leanok
  \uses{thm:transformerBlock_has_vjp_mat, thm:identityMat_has_vjp, thm:vjpMat_comp}
  Covers ViT-Tiny (12 blocks), ViT-Base (12), ViT-Large (24).
\end{theorem}

\begin{theorem}[ViT body: the grand finale]
  \label{thm:vit_body_has_vjp_mat}
  \lean{Proofs.vit_body_has_vjp_mat}
  \leanok
  \uses{thm:transformerTower_has_vjp_mat, thm:layerNorm_per_token_has_vjp_mat, thm:vjpMat_comp}
  The full ViT transformer backbone as one \(\HasVJPMat\), derived from
  exactly one new axiom (\texttt{mhsa\_has\_vjp\_mat}).
\end{theorem}

\chapter{Bestiary of Architectures}
\label{chap:bestiary}

The proof chapters above cover every layer shipped in the Phase 1--3 trainers
(MLP, CNN, ResNet, MobileNet, EfficientNet, ViT). The \textbf{bestiary} is
the book's Part 2: a catalogue of famous architectures expressed as pure
\texttt{NetSpec} values --- no training runs, no VJP commitments, just
``here's what this architecture looks like in $\sim$15 lines of Lean.''

Each bestiary entry introduces zero or one new \texttt{Layer} constructor
for its architectural idiom. The primitive is bundled at the right
abstraction level (a single Transformer encoder block, a whole Mamba
block, a stage of Swin attention, etc.) --- formalizing every sub-op of
a modern architecture would be an asymptote we don't approach.

\section{Bestiary-only \texttt{Layer} primitives}
\label{sec:bestiary_primitives}

Ten new \texttt{Layer} constructors were added by the bestiary chapters.
None are codegen-backed (MlirCodegen emits \texttt{// UNSUPPORTED}); the
goal is pedagogical shape + parameter accounting, not training runs.

\begin{definition}[Mamba block]
  \label{layer:mambaBlock}
  \lean{Layer.mambaBlock}
  \leanok
  Selective state-space block in the S6 formulation (Gu \& Dao 2023).
  Signature: \texttt{mambaBlock (dim stateSize expand nBlocks : Nat)}.
  Bundles RMSNorm + linear expand + depthwise 1D conv + SiLU +
  selective-scan SSM + gated product + output projection, for
  \texttt{nBlocks} stacked layers. The selective scan is the novel
  primitive; everything else could be decomposed to existing layers if
  we cared to unpack the bundle.
\end{definition}

\begin{definition}[Swin Transformer stage]
  \label{layer:swinStage}
  \lean{Layer.swinStage}
  \leanok
  Windowed multi-head self-attention at fixed spatial resolution (Liu
  et al.\ 2021). Signature:
  \texttt{swinStage (dim heads mlpDim windowSize nBlocks : Nat)}. Internal
  blocks alternate W-MSA and SW-MSA (shifted-window) to let information
  cross window boundaries.
\end{definition}

\begin{definition}[Patch merging]
  \label{layer:patchMerging}
  \lean{Layer.patchMerging}
  \leanok
  Swin's 2$\times$2 spatial downsample + linear channel projection
  (\texttt{inDim $\to$ outDim}). Transformer-side analog of a stride-2
  conv.
\end{definition}

\begin{definition}[UNet encoder stage]
  \label{layer:unetDown}
  \lean{Layer.unetDown}
  \leanok
  $2 \times$ \texttt{(conv3$\times$3 + BN + ReLU)} then \texttt{maxPool-2}.
  Saves its pre-pool activation as a skip for the matching
  \texttt{unetUp}. Signature: \texttt{unetDown (ic oc : Nat)}.
\end{definition}

\begin{definition}[UNet decoder stage]
  \label{layer:unetUp}
  \lean{Layer.unetUp}
  \leanok
  Transposed-conv $2\times$ upsample + concat with matching skip +
  $2 \times$ \texttt{(conv3$\times$3 + BN + ReLU)}. Signature:
  \texttt{unetUp (ic oc : Nat)}, where \texttt{oc} is both the output
  channel count and the expected skip width.
\end{definition}

\begin{definition}[Transformer decoder (DETR-style)]
  \label{layer:transformerDecoder}
  \lean{Layer.transformerDecoder}
  \leanok
  \texttt{nBlocks} blocks of self-attention over
  \texttt{nQueries} learned object queries + cross-attention against an
  encoder output + FFN. The query embedding is part of the layer's
  parameters. Signature:
  \texttt{transformerDecoder (dim heads mlpDim nBlocks nQueries : Nat)}.
\end{definition}

\begin{definition}[DETR prediction heads]
  \label{layer:detrHeads}
  \lean{Layer.detrHeads}
  \leanok
  Per-query class head (linear \texttt{dim $\to$ nClasses+1} with the
  ``no object'' slot) + box head (3-layer MLP to 4 scalars
  \texttt{(cx, cy, w, h)}). Signature: \texttt{detrHeads (dim nClasses : Nat)}.
\end{definition}

\begin{definition}[ShuffleNet v1 stage]
  \label{layer:shuffleBlock}
  \lean{Layer.shuffleBlock}
  \leanok
  Grouped 1$\times$1 conv + channel-shuffle permutation + 3$\times$3
  depthwise + grouped 1$\times$1 conv, for \texttt{nUnits} units
  (first downsampling, rest residual). The shuffle is parameter-free;
  grouping reduces 1$\times$1 cost by $g$. Signature:
  \texttt{shuffleBlock (ic oc groups nUnits : Nat)}.
\end{definition}

\begin{definition}[ShuffleNet v2 stage]
  \label{layer:shuffleV2Block}
  \lean{Layer.shuffleV2Block}
  \leanok
  \texttt{nUnits} v2 units (Ma et al.\ 2018). Basic unit (stride 1):
  channel-split $[X_1, X_2]$, leave $X_1$ alone, run $X_2$ through
  $1\times 1 \to 3\times 3$ DW $\to 1\times 1$ at half-width, concat, then
  channel-shuffle. Downsample unit (stride 2): both branches see the
  full input; left does DW-3$\times$3-stride-2 $+$ 1$\times$1, right does
  1$\times$1 $+$ DW-3$\times$3-stride-2 $+$ 1$\times$1; concat doubles
  channels. v2 throws out v1's grouped 1$\times$1 convs (G2) and
  skip-add (G4) per the paper's practical guidelines. Signature:
  \texttt{shuffleV2Block (ic oc nUnits : Nat)}.
\end{definition}

\begin{definition}[Evoformer block (AlphaFold 2)]
  \label{layer:evoformerBlock}
  \lean{Layer.evoformerBlock}
  \leanok
  Dual-representation (MSA + pair) joint update: MSA row-attention with
  pair bias + MSA column-attention + MSA transition +
  outer-product-mean ($\to$ pair) + triangle multiplicative (outgoing,
  incoming) + triangle self-attention (starting node, ending node) +
  pair transition. Signature:
  \texttt{evoformerBlock (msaChannels pairChannels nBlocks : Nat)}.
  The triangulation-aware operations are the key inductive bias.
\end{definition}

\begin{definition}[Structure Module (AlphaFold 2)]
  \label{layer:structureModule}
  \lean{Layer.structureModule}
  \leanok
  Recurrent Invariant Point Attention + backbone frame update +
  side-chain $\chi$-angle prediction. \textbf{Weights shared across
  \texttt{nBlocks} rounds} --- param count does not multiply by
  \texttt{nBlocks}. Signature:
  \texttt{structureModule (singleChannels pairChannels nBlocks : Nat)}.
\end{definition}

\begin{definition}[MobileViT block]
  \label{layer:mobileVitBlock}
  \lean{Layer.mobileVitBlock}
  \leanok
  Hybrid local-conv + patch-level transformer (Mehta \& Rastegari
  2022). Local 3$\times$3 conv $\to$ 1$\times$1 projection to transformer
  dim $\to$ unfold into patches $\to$ $L$ transformer blocks across
  patches $\to$ fold back $\to$ 1$\times$1 projection back $\to$ concat
  with input $\to$ 3$\times$3 fusion. Signature:
  \texttt{mobileVitBlock (ic dim heads mlpDim nTxBlocks : Nat)}.
  The unfold/fold operations are \texttt{pdiv\_reindex}-style shape
  transformations; all the genuinely new math is already covered by
  the transformer proof chapter.
\end{definition}

\begin{definition}[ConvNeXt stage]
  \label{layer:convNextStage}
  \lean{Layer.convNextStage}
  \leanok
  ConvNeXt residual block $\times$ \texttt{nBlocks}: 7$\times$7 depthwise
  conv $\to$ LayerNorm $\to$ 1$\times$1 expand (to $4c$) $\to$ GELU $\to$
  1$\times$1 project (back to $c$) + residual. The transformer-era CNN
  block (Liu et al.\ 2022). Does not include downsampling; see
  \texttt{convNextDownsample}. Signature:
  \texttt{convNextStage (channels nBlocks : Nat)}.
\end{definition}

\begin{definition}[ConvNeXt downsampling]
  \label{layer:convNextDownsample}
  \lean{Layer.convNextDownsample}
  \leanok
  Inter-stage spatial halving $+$ channel-doubling: LayerNorm $+$
  2$\times$2 conv stride 2. Swin's \texttt{patchMerging} for CNNs.
  Signature: \texttt{convNextDownsample (ic oc : Nat)}.
\end{definition}

\begin{definition}[WaveNet block]
  \label{layer:waveNetBlock}
  \lean{Layer.waveNetBlock}
  \leanok
  One stack of \texttt{nLayers} dilated causal residual blocks with
  doubling dilation rates $2^0, 2^1, \ldots, 2^{\texttt{nLayers}-1}$
  (van den Oord et al.\ 2016). Each block: dilated 2-tap causal conv
  $\to$ gated activation $\tanh(\text{filter}) \odot \sigma(\text{gate})$
  $\to$ 1$\times$1 project back to \texttt{residualCh} (residual path) +
  1$\times$1 skip projection to \texttt{skipCh}. Skip outputs across
  blocks are summed into the final head. Signature:
  \texttt{waveNetBlock (residualCh skipCh nLayers : Nat)}. Output
  channels are \texttt{skipCh}: bestiary convention picks the skip path
  as the "forward" output since it's what feeds the final classifier.
\end{definition}

\begin{definition}[Positional encoding]
  \label{layer:positionalEncoding}
  \lean{Layer.positionalEncoding}
  \leanok
  Sinusoidal frequency basis (Vaswani 2017, reused by NeRF 2020):
  $\gamma(p) = (\sin 2^0 \pi p, \cos 2^0 \pi p, \ldots, \sin 2^{L-1} \pi p, \cos 2^{L-1} \pi p)$.
  Zero trainable parameters --- it's a deterministic lift of a
  low-dim coordinate into a high-frequency feature space where an
  MLP has enough wiggle room to represent sharp details. Output dim
  $=$ inputDim $\cdot 2 \cdot$ numFrequencies. Signature:
  \texttt{positionalEncoding (inputDim numFrequencies : Nat)}.
\end{definition}

\begin{definition}[NeRF MLP core]
  \label{layer:nerfMLP}
  \lean{Layer.nerfMLP}
  \leanok
  The whole NeRF network (Mildenhall et al.\ 2020) bundled as one
  primitive: 8 hidden ReLU-FC layers of \texttt{hiddenDim}, mid-skip
  concatenating $\gamma(x)$ at layer 5, dual output heads
  (1-dim volume density $\sigma$ + 3-dim RGB via a direction-conditioned
  branch). Under 600K parameters at the canonical config. Signature:
  \texttt{nerfMLP (encodedPosDim encodedDirDim hiddenDim : Nat)}.
\end{definition}

\begin{definition}[Darknet residual block]
  \label{layer:darknetBlock}
  \lean{Layer.darknetBlock}
  \leanok
  YOLOv3's Darknet-53 residual stack (Redmon \& Farhadi 2018).
  \texttt{nBlocks} residual blocks at fixed \texttt{channels}, each
  being 1$\times$1 conv ($c \to c/2$) $+$ 3$\times$3 conv ($c/2 \to c$)
  $+$ residual add. Lighter than a standard ResNet bottleneck; heavier
  than a ResNet-18 basic block. Signature:
  \texttt{darknetBlock (channels nBlocks : Nat)}.
\end{definition}

\begin{definition}[Cross-Stage Partial block]
  \label{layer:cspBlock}
  \lean{Layer.cspBlock}
  \leanok
  CSP (Wang et al.\ 2019), used by YOLOv4 onward. Splits input into two
  halves, processes one half through a stack of residual blocks, then
  concatenates with the untouched half and 1$\times$1-projects to
  \texttt{oc}. The specific inner block varies across YOLO versions (C3
  in v5, C2f in v8, C3k2 in v11); this primitive approximates all three
  at the same abstraction level. Signature:
  \texttt{cspBlock (ic oc nBlocks : Nat)}.
\end{definition}

\begin{definition}[Inception module]
  \label{layer:inceptionModule}
  \lean{Layer.inceptionModule}
  \leanok
  The GoogLeNet parallel-branch module (Szegedy et al.\ 2014). Four
  branches computed in parallel: 1$\times$1 conv (b1 channels); 1$\times$1
  reduce then 3$\times$3 (b2); 1$\times$1 reduce then 5$\times$5 (b3);
  3$\times$3 maxPool then 1$\times$1 (b4). Concat along channels for
  \texttt{b1 + b2 + b3 + b4} outputs. The 1$\times$1 dimension reducers on
  branches 2 and 3 are the paper's trick --- they make the expensive
  3$\times$3 and 5$\times$5 convs operate on reduced channel counts.
  Signature: \texttt{inceptionModule (ic b1 b2reduce b2 b3reduce b3 b4 : Nat)}.
\end{definition}

\section{Bestiary entries}
\label{sec:bestiary_entries}

The entries are grouped by task domain. The first block (\emph{vision
classifiers}) is where Part 1's VJP'd primitives show up at real-world
scale --- every layer is one you've already seen proved correct.
Subsequent blocks step out to detection, segmentation, reinforcement
learning, and the non-vision outliers (language, audio, 3D,
multimodal, science).

\subsection*{\textmd{\textsc{Group I.}} Vision classifiers --- Part 1's primitives at scale}

Image-classification backbones built out of conv / pool / batch-norm /
residual / attention / patch-embed --- the exact layer kit VJP'd in
Part 1. If a chapter of Part 1 proved it, a bestiary entry below puts
it to work.

\subsection*{LeNet (\texttt{Bestiary/LeNet.lean})}
The 1998 original. Zero new primitives --- just \texttt{conv2d}
$+$ \texttt{maxPool} $+$ \texttt{dense}. Variants: LeNet-5 (61K params,
the canonical CNN) and LeNet-300-100 (266K params, pure-MLP baseline).
Historical importance; still the ground-truth pattern every later CNN
riffs on.

\subsection*{AlexNet (\texttt{Bestiary/AlexNet.lean})}
Zero new primitives --- five convs, three FCs, pools. The 2012
ImageNet winner that restarted modern deep learning. Variants:
AlexNet (62M, paper-exact at 60M) + tiny CIFAR fixture. $\sim$58M of
the 62M live in the three FC layers, which is precisely why every
post-2015 CNN dropped FC stacks for \texttt{globalAvgPool} + one final
dense. LRN is omitted (replaced in the field by BatchNorm in 2015);
dropout is training-time only.

\subsection*{SqueezeNet (\texttt{Bestiary/SqueezeNet.lean})}
\emph{Zero} new primitives --- the \texttt{.fireModule} constructor was
already in \texttt{Types.lean}. AlexNet-level accuracy in 1.25M params
via the fire module: squeeze 1$\times$1 conv followed by parallel
expand 1$\times$1 + 3$\times$3 convs concatenated. The early
efficiency-CNN family alongside MobileNet and ShuffleNet. Variants:
SqueezeNet 1.0 (1.25M, paper-exact), 1.1 (1.24M --- earlier downsample),
tiny fixture.

\subsection*{Inception v1 / v3 / v4 (\texttt{Bestiary/Inception.lean})}
Uses \S~\ref{layer:inceptionModule}. GoogLeNet's parallel multi-scale
conv-branches concatenated along channels; the 1$\times$1 dimension
reducer was invented here. Variants: GoogLeNet (7M, paper-exact),
Inception-v3 (23.8M, paper 23M), Inception-v4 (33M, paper 42M ---
a bit low because our unified module approximates v3/v4's richer
module catalog). Auxiliary classifiers (v1) and asymmetric
factorizations (v3/v4) are bestiary omissions.

\subsection*{Xception (\texttt{Bestiary/Xception.lean})}
Zero new primitives --- reuses existing \texttt{.separableConv}.
``Extreme Inception'': every conv is a depthwise-separable conv. The
design choice that made MobileNet possible a year later. Variants:
Xception (21.9M, paper-exact at 22M), tiny fixture. Residual skips
around each block-of-three-sep-convs are implicit in the linear
NetSpec.

\subsection*{ShuffleNet (\texttt{Bestiary/ShuffleNet.lean})}
Uses \S~\ref{layer:shuffleBlock}. Variants: ShuffleNet 0.5$\times$ /
1.0$\times$ / 2.0$\times$ at $g=3$, plus a tiny fixture.

\subsection*{ShuffleNet v2 (\texttt{Bestiary/ShuffleNetV2.lean})}
Uses \S~\ref{layer:shuffleV2Block}. The efficient-CNN paper that
called out FLOPs as a bad latency proxy: measured memory-access cost
(MAC) directly and derived four practical guidelines (equal channel
widths, avoid grouped convs, avoid fragmentation, avoid element-wise
ops). v2's architecture throws out everything in v1 that violated
those rules --- no grouped 1$\times$1 convs, no skip-add --- in
favor of channel-split + identity + concat + channel-shuffle.
Variants: 0.5$\times$ (1.37M, paper 1.4M), 1.0$\times$ (2.29M, paper
2.3M), 1.5$\times$ (3.52M, paper 3.5M), 2.0$\times$ (7.41M,
paper-exact), tiny fixture. All widths within 2\% of paper.

\subsection*{MobileViT (\texttt{Bestiary/MobileViT.lean})}
Uses \S~\ref{layer:mobileVitBlock} plus existing
\texttt{invertedResidual} for the MV2 stages. Hybrid mobile backbone:
MobileNet V2 body with MobileViT blocks replacing some of the deeper
inverted-residual stages. Variants: MobileViT-S (5.6M, paper-exact),
XS (2.3M), XXS (1.3M), tiny fixture.

\subsection*{ConvNeXt (\texttt{Bestiary/ConvNeXt.lean})}
Uses \S~\ref{layer:convNextStage} and \S~\ref{layer:convNextDownsample}.
ResNet-50 modernized with transformer design choices: 7$\times$7
depthwise + inverted bottleneck + LN + GELU + separate inter-stage
downsamples. Proof-by-construction that CNNs didn't need to die in
2021. Variants: T (28M), S (50M), B (89M), L (198M), tiny fixture ---
all matching the paper within 1\%.

\subsection*{Swin Transformer (\texttt{Bestiary/SwinT.lean})}
Uses \S~\ref{layer:swinStage} and \S~\ref{layer:patchMerging}. Variants:
Swin-T / -S / -B / tiny. Swin-T lands at 28M params matching the paper
exactly.

\subsection*{\textmd{\textsc{Group II.}} Object detection}

Localize and classify. Detection heads are where the linear
\texttt{NetSpec} shape starts to creak: multi-scale FPN outputs need a
graph, not a list. The bestiary entries show the single-scale view and
defer the multi-head refactor to the limitations discussion.

\subsection*{YOLO v1/v3/v5/v8/v11 (\texttt{Bestiary/YOLO.lean})}
v1 uses zero new primitives (\texttt{conv2d} + \texttt{maxPool} +
\texttt{flatten} + \texttt{dense}; the YOLO-ness lives in the loss and
output reshape, not the architecture). v3 adds \S~\ref{layer:darknetBlock}
for the Darknet-53 body. v5/v8/v11 use \S~\ref{layer:cspBlock} for
their CSPDarknet backbones. Variants: YOLOv1 (271M, paper-exact), fast,
tiny, YOLOv3 (40M backbone), YOLOv5s/m (3M/8M single-scale), YOLOv8n/s
(0.7M/2.9M single-scale), YOLOv11n/m (backbone only). Multi-scale FPN
detection heads don't linearize --- same skip-connection problem as
UNet; all entries show a single-scale view.

\subsection*{DETR (\texttt{Bestiary/DETR.lean})}
Uses \S~\ref{layer:transformerDecoder} and \S~\ref{layer:detrHeads},
plus existing \texttt{bottleneckBlock} (ResNet backbone),
\texttt{patchEmbed} (absorbs the DETR 1$\times$1 channel reduce), and
\texttt{transformerEncoder}. Variants: DETR-R50 (41M, paper-exact),
DETR-R101 (60M), tiny.

\subsection*{\textmd{\textsc{Group III.}} Semantic segmentation}

Pixel-level labeling. Symmetric encoder/decoder with skip connections
is the recurring pattern; UNet is the canonical instance, and the same
shape later became the diffusion-model backbone.

\subsection*{UNet (\texttt{Bestiary/UNet.lean})}
Uses \S~\ref{layer:unetDown} and \S~\ref{layer:unetUp}. Skip
connections are implicit --- the $i$-th \texttt{unetUp} from the bottom
pairs with the $i$-th \texttt{unetDown} from the top. Variants:
original (1-channel, 2-class), RGB, small, tiny. Original lands at 31M
params matching Ronneberger.

\subsection*{\textmd{\textsc{Group IV.}} Reinforcement learning}

Two-headed (policy + value) networks wrapped in a self-play + MCTS
outer loop. The architectural side is a stack of residual CNN blocks;
the complexity lives in the outer loop, not the network.

\subsection*{AlphaZero (\texttt{Bestiary/AlphaZero.lean})}
Uses existing primitives only (\texttt{convBn} + \texttt{residualBlock}
+ \texttt{conv2d} + \texttt{dense}). Two-headed (policy + value),
expressed as two separate \texttt{NetSpec} values sharing the body in
prose. Variants: AlphaGo Zero (Go), AlphaZero chess, tiny fixture.

\subsection*{MuZero (\texttt{Bestiary/MuZero.lean})}
Zero new primitives --- three ResNet-style networks (representation,
dynamics, prediction) reusing \texttt{convBn} + \texttt{residualBlock}
+ \texttt{dense}. The architectural novelty is the three-network
factoring, not any single layer type. Five \texttt{NetSpec} values per
variant (rep, dyn body, dyn reward head, pred policy, pred value).
Variants: Go, Atari (representation only), tiny.

\subsection*{\textmd{\textsc{Group V.}} Beyond vision}

Architectures where the task domain is not a 2D image: language,
audio, 3D scene reconstruction, multimodal embedding, scientific.
Several of these (NeRF, CLIP) have essentially \emph{no} architectural
novelty --- the interesting work lives in the data, loss, or training
procedure, and the bestiary entry exists to make that point.

\subsection*{Mamba (\texttt{Bestiary/Mamba.lean})}
Uses \S~\ref{layer:mambaBlock}. Variants: Mamba-130M / 370M / 790M /
tiny, matching Gu \& Dao's param counts within $\sim$5\%.

\subsection*{BERT / RoBERTa (\texttt{Bestiary/BERT.lean})}
Zero new primitives --- uses \texttt{.transformerEncoder}, same kit as
ViT / DETR, plus \texttt{.dense vocab$\to$dim} standing in for the
token-embedding table (faithful param count, shape semantics cheat
since a linear \texttt{NetSpec} can't express the $L \to L \times D$
lookup). Variants: BERT-base (109M, paper 110M), BERT-large (335M,
paper 340M), RoBERTa-base (124M, paper 125M), RoBERTa-large (355M,
paper-exact), tinyBERT fixture. The architectural lesson is that
RoBERTa = BERT; all RoBERTa gains came from training procedure
(dynamic masking, more data, bigger batches, dropped NSP) --- none
of which lives in the NetSpec.

\subsection*{WaveNet (\texttt{Bestiary/WaveNet.lean})}
Uses \S~\ref{layer:waveNetBlock}. Dilated causal convolutions for raw
audio sample prediction: exponential receptive field, linear parameter
growth. The foundation of neural TTS and PixelCNN. Variants: single
stack (0.4M), 3-stack (paper setup, with a NetSpec-linearity
approximation), music (single stack, 4.1M), tiny. One honest
limitation: the residual-vs-skip dual-output pattern doesn't
linearize cleanly, so the 3-stack variant uses a simplified
channel-flow approximation.

\subsection*{NeRF (\texttt{Bestiary/NeRF.lean})}
Uses \S~\ref{layer:positionalEncoding} and \S~\ref{layer:nerfMLP}. The
"it's literally just an MLP" paper. Under 600K params at canonical
config; the architectural novelty is nonexistent. What makes NeRF work
is the positional encoding + the volumetric-rendering loss ---
both \emph{outside} the network, not layers in it. Variants:
canonical (593K), fast (167K hidden=128), tiny fixture.

\subsection*{CLIP (\texttt{Bestiary/CLIP.lean})}
\emph{Zero} new primitives. Two textbook encoders glued together by a
contrastive loss: a ResNet-50 or ViT-B for images, a 12-layer causal
transformer for text, each with a final linear projection to a shared
embedding space. Variants: CLIP-RN50, CLIP-ViT-B/32 (151M, paper-exact),
CLIP-ViT-L/14 (427M, paper-exact), tiny fixture. The architectural
lesson from CLIP is identical to NeRF's: the novelty lives in data and
objective, not in layer design.

\subsection*{AlphaFold 2 Evoformer (\texttt{Bestiary/Evoformer.lean})}
Uses \S~\ref{layer:evoformerBlock} and \S~\ref{layer:structureModule}.
Dual-representation (MSA + pair) doesn't fit a linear \texttt{NetSpec}
cleanly; the bestiary shows the two bundled primitives and notes the
limitation. Variants: full (76M), mini, tiny.

\appendix

\chapter{Axioms at a glance}
\label{app:axioms}

The Verified VJP Proofs suite contains 30 axioms and 48 theorems across
nine Lean files, with zero \texttt{sorry}s. The axioms fall into three
buckets:

\begin{itemize}
  \item \textbf{Calculus primitives} (8): chain rule, linearity, product
    rule, identity, reindex --- elementary facts about the partial-derivative
    function \(\pdiv\). Correspond directly to Mathlib \texttt{HasDerivAt}
    / \texttt{HasFDerivAt} lemmas.
  \item \textbf{Elementary Jacobian facts} (about 11): the Jacobian of a
    specific operation (ReLU diagonal, softmax rank-1, layer-norm 3-term,
    etc.), stated in \(\pdiv\)-land rather than bridging to Mathlib.
  \item \textbf{Bundled layer VJPs} (about 11): each opaque forward
    (\texttt{conv2d}, \texttt{depthwiseConv2d}, \texttt{mhsa\_layer}) paired
    with a bundled \(\HasVJP\) existence claim. Gradient-checked numerically
    in \texttt{check\_axioms.py}.
\end{itemize}

Everything else --- 48 theorems --- is composition over these axioms, glued
by the three structural rules (chain, fan-in, product) and five
closed-form Jacobian tricks (diagonal, sparse toeplitz, binary selection,
rank-1 correction, outer product).
